---
title: "`r getOption('irxclean.report_env')$report_title`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    self_contained: true
---

<style>
/* InsightRX brand styles */
body {
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto,
    "Helvetica Neue", Arial, sans-serif;
  color: #111111;
}
h1, h2, h3, h4 {
  color: #003B4F;
}
a {
  color: #00769E;
}
a:hover {
  color: #20794D;
}
/* TOC */
.tocify .tocify-item a, .tocify .tocify-subheader .tocify-item a {
  color: #003B4F;
}
.tocify-subheader .tocify-item a {
  color: #5E5E5E;
}
/* Table */
table th {
  background-color: #003B4F;
  color: white;
}
table tr:nth-child(even) {
  background-color: #F0F0F0;
}
/* Section dividers */
hr {
  border-color: #003B4F;
  opacity: 0.2;
}
/* Title banner */
.fluid-row:first-of-type {
  background-color: #003B4F;
  color: white;
  padding: 20px;
  margin-bottom: 20px;
}
/* Collapsible details block */
details {
  border: 1px solid #ddd;
  border-radius: 4px;
  padding: 0.5em 1em;
  margin-bottom: 1.5em;
}
details summary {
  font-weight: bold;
  color: #003B4F;
  cursor: pointer;
  padding: 0.3em 0;
}
details summary:hover {
  color: #00769E;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5
)

# Retrieve data objects from the options environment
.env <- getOption("irxclean.report_env")
results                 <- .env$results
data                    <- .env$data
prop_error_param        <- .env$prop_error_param
add_error_param         <- .env$add_error_param
continuous_cov_cols     <- .env$continuous_cov_cols
categorical_cov_cols    <- .env$categorical_cov_cols
id_col                  <- .env$id_col
time_col                <- .env$time_col
evid_col                <- .env$evid_col
tad_col                 <- .env$tad_col
tad_bin_width           <- .env$tad_bin_width
stability_threshold_pct <- .env$stability_threshold_pct

library(irxclean)
```

---

# Overview

```{r overview}
n_iter    <- max(results$phi$ITERATION)
n_removed <- nrow(results$rem)
rem_ids   <- unique(results$rem[[id_col]])
```

This report summarises the iterative detection analysis performed with
**irxclean**, identifying potentially erroneous observations for user
investigation and review.  Final exclusion decisions rest with the analyst.

| | |
|---|---|
| Iterations completed | `r n_iter` |
| Observations removed | `r n_removed` |
| Subjects affected | `r length(rem_ids)` |

---

# Key Results

## Observations Removed

```{r removed-table}
knitr::kable(
  results$rem[, intersect(
    c(id_col, "TIME", "DV", "CWRES", "ITERATION"), names(results$rem)
  )],
  digits  = 3,
  caption = "Observations removed in order of removal iteration"
)
```

## pOFV Change

```{r pofv-plot, fig.cap="Change in pseudo-OFV (pOFV) at each iteration, restricted to subjects with no observations removed. Each point is the pOFV improvement attributable to removing that iteration's observation. Three log-linear fits (y ~ log(x), 95% CI) are overlaid: using 100%, first 50%, and first 25% of removals."}
plot_removal_metrics(results, metric = "pOFV", verbose = FALSE)
```

```{r prop-error-section, results='asis', echo=FALSE}
if (!is.null(prop_error_param) && prop_error_param %in% names(results$par)) {
  cat("## Proportional Error Stability\n\n")
}
```

```{r prop-error-plot, eval=!is.null(prop_error_param) && !is.null(results$par) && prop_error_param %in% names(results$par), fig.cap="Proportional residual error parameter stability across removal iterations."}
irxclean:::.plot_param_trajectory(
  results,
  prop_error_param,
  title = "Proportional residual error across iterations"
)
```

```{r add-error-section, results='asis', echo=FALSE}
if (!is.null(add_error_param) && add_error_param %in% names(results$par)) {
  cat("## Additive Error Stability\n\n")
}
```

```{r add-error-plot, eval=!is.null(add_error_param) && !is.null(results$par) && add_error_param %in% names(results$par), fig.cap="Additive residual error parameter stability across removal iterations."}
irxclean:::.plot_param_trajectory(
  results,
  add_error_param,
  title = "Additive residual error across iterations"
)
```

## nRMSE Trend

```{r rmse-plot, fig.cap="Normalised RMSE (nRMSE = RMSE / mean(DV)) across iterations. A decreasing trend indicates progressive improvement in predictive accuracy."}
plot_removal_metrics(results, metric = "rmse", verbose = FALSE)
```

---

```{r open-details, results='asis', echo=FALSE}
cat('\n<details>\n<summary><strong>Detailed Results</strong> (click to expand)</summary>\n')
```

### THETA Stability {.unlisted .unnumbered}

```{r theta-plot, fig.cap="THETA parameter estimates normalised to final iteration value. Fixed parameters excluded."}
plot_removal_metrics(results, metric = "thetas", verbose = FALSE)
```

### OMEGA Stability {.unlisted .unnumbered}

```{r omega-plot, fig.cap="OMEGA diagonal variance estimates normalised to final iteration value. Fixed parameters and BLOCK SAME entries excluded."}
plot_removal_metrics(results, metric = "omegas", verbose = FALSE)
```

```{r sigma-section, results='asis', echo=FALSE}
# Show SIGMA section only when at least one SIGMA is non-fixed
has_free_sigma <- {
  sigma_cols <- grep("^SIGMA", names(results$par), value = TRUE)
  if (length(sigma_cols) > 0) {
    any(vapply(results$par[, sigma_cols, drop = FALSE], function(x) {
      v <- x[!is.na(x)]
      length(v) > 0L && !all(v == v[1L])
    }, logical(1L)))
  } else {
    FALSE
  }
}
if (has_free_sigma) cat("### SIGMA Stability {.unlisted .unnumbered}\n\n")
```

```{r sigma-plot, eval=has_free_sigma, fig.cap="SIGMA diagonal variance estimates normalised to final iteration value. Fixed parameters excluded."}
plot_removal_metrics(results, metric = "sigmas", verbose = FALSE)
```

### Model Stability Assessment {.unlisted .unnumbered}

```{r stability}
stability <- check_model_stability(
  results,
  threshold_pct = stability_threshold_pct,
  verbose       = FALSE
)
```

**Threshold:** `r stability_threshold_pct`%
**RMSE trend:** `r stability$rmse_trend`
**Unstable parameters:** `r stability$n_flagged_params`

Each free parameter is classified using three metrics computed across all
removal iterations:

| Metric | Definition |
|---|---|
| RSE (%) | SD / \|mean\| x 100 -- overall variability |
| Reversals | Sign changes in consecutive differences -- measures trend bouncing |
| CV late (%) | CV of the final third of iterations -- measures whether the parameter has settled |

| Status | Criteria |
|---|---|
| **unstable** | RSE > 50% AND reversals >= 3, OR CV late > 10% |
| **drifted** | Drift (first to last) > threshold / 2, but settled monotonically |
| **stable** | Neither of the above |

A *drifted* parameter is expected behaviour when a genuine outlier is removed;
an *unstable* parameter bounces or has not settled by the final iteration and
warrants investigation.

```{r stability-flags}
flag_tbl <- stability_flag_table(stability)
if (nrow(flag_tbl) > 0) {
  knitr::kable(flag_tbl, digits = 2,
               caption = "Parameters that did not reach a stable estimate")
} else {
  cat("All parameters reached a stable estimate across iterations.\n")
}
```

```{r stability-plots}
pl <- plot(stability)
```

```{r stability-trajectory, eval=stability$n_flagged_params > 0, fig.cap="Parameter trajectories (% change from baseline) coloured by stability status. Unstable = bouncing/unsettled; drifted = shifted monotonically to a new value; stable = low variability."}
if (!is.null(pl$parameter_drift)) print(pl$parameter_drift)
```

```{r stability-rse, fig.cap="RSE (SD / |mean| x 100%) per parameter across iterations. Red dashed line = stability threshold."}
if (!is.null(pl$param_rse)) print(pl$param_rse)
```

```{r close-details, results='asis', echo=FALSE}
cat('\n</details>\n')
```

---

```{r demo-header, results='asis', echo=FALSE}
has_demo <- !is.null(data) &&
  (!is.null(continuous_cov_cols) || !is.null(categorical_cov_cols))
if (has_demo) {
  cat("# Demographic Comparison\n\n")
  cat("Subjects with removed observations are compared to the full dataset ")
  cat("to screen for demographic bias in the removal algorithm. ")
  cat("Significant covariates are highlighted with **red bold** titles.\n\n")
}
```

```{r demographic-data, eval=has_demo, include=FALSE}
demo <- plot_demographic_comparison(
  data             = data,
  results          = results,
  continuous_cols  = continuous_cov_cols,
  categorical_cols = categorical_cov_cols,
  id_col           = id_col,
  time_col         = time_col,
  evid_col         = evid_col,
  tad_col          = tad_col,
  tad_bin_width    = tad_bin_width
)

# Pre-render panel chunks for tabbed output when >1 panel
demo_panel_chunks <- if (length(demo$panels) > 1) {
  lapply(seq_along(demo$panels), function(i) {
    knitr::knit_child(
      text = c(
        sprintf("### Panel %d", i),
        "",
        sprintf(
          "```{r demo-cov-panel-%d, echo=FALSE, fig.width=12, fig.height=9}",
          i
        ),
        sprintf("demo$panels[[%d]]", i),
        "```",
        ""
      ),
      envir = environment(),
      quiet = TRUE
    )
  })
} else {
  NULL
}
```

```{r demo-stats, eval=has_demo, results='asis', echo=FALSE}
cat("## Statistical Tests\n\n")
cat(
  knitr::knit_child(
    text  = c("```{r, echo=FALSE}", "knitr::kable(demo$stats, digits=3)", "```"),
    envir = environment(),
    quiet = TRUE
  ),
  sep = "\n"
)
```

```{r demo-covariates, eval=has_demo, results='asis', echo=FALSE}
if (length(demo$panels) > 1) {
  cat("## Covariate Distributions {.tabset}\n\n")
  cat(unlist(demo_panel_chunks), sep = "\n")
} else if (length(demo$panels) == 1) {
  cat("## Covariate Distributions\n\n")
  cat(
    knitr::knit_child(
      text = c(
        "```{r demo-cov-single, echo=FALSE, fig.width=12, fig.height=9}",
        "demo$panels[[1]]",
        "```"
      ),
      envir = environment(),
      quiet = TRUE
    ),
    sep = "\n"
  )
}
```

```{r demo-tad, eval=has_demo && !is.null(demo$tad_plot), results='asis',
    echo=FALSE}
cat("## Removal Position on PK Curve\n\n")
cat("*Proportion of observations removed per time-after-dose bin.
Red bars are significantly elevated relative to the overall removal rate
(one-sided binomial test, Bonferroni-corrected across bins).*\n\n")
cat(
  knitr::knit_child(
    text  = c(
      "```{r demo-tad-plot, echo=FALSE, fig.width=10, fig.height=4}",
      "demo$tad_plot",
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ),
  sep = "\n"
)
```

---

# Session Information

```{r session-info}
si <- utils::sessionInfo()
cat(sprintf("irxclean version: %s\n", utils::packageVersion("irxclean")))
cat(sprintf("R version: %s\n", R.version$version.string))
cat(sprintf("Date: %s\n", Sys.time()))
```
