---
title: "Simulated Busulfan Dataset: End-to-End irxclean Workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulated Busulfan Dataset: End-to-End irxclean Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-model, include=FALSE, eval=FALSE}
# Install bundled busulfan model package if not already present
if (!requireNamespace("pkbusulfanucsf2025", quietly = TRUE)) {
  PKPDsim::model_from_api(
    model      = "pk_busulfan_ucsf_2025",
    url        = system.file("models", "pk_busulfan_ucsf_2025.json5",
                             package = "irxclean"),
    to_package = TRUE,
    force      = TRUE,
    install    = TRUE,
    verbose    = FALSE
  )
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5
)
library(irxclean)
```

## Overview

This vignette walks through a complete `irxclean` analysis using a **simulated
busulfan pharmacokinetic dataset**.  Busulfan is a narrow-therapeutic-index
alkylating agent used in haematopoietic stem-cell transplant conditioning;
individualised dosing based on therapeutic drug monitoring (TDM) is standard
of care.

The iterative-flagging stage (Section 3) runs NONMEM when PsN is available and
caches results locally; subsequent renders load the cache.  If neither is
present, a 5-iteration pre-computed result bundled in `inst/testdata/` is used
as a fallback.  All other steps run live against the simulated dataset.

### Dataset description

The dataset was generated with **PKPDsim** using the `pkbusulfanucsf2025` model
package, a two-compartment IV busulfan population PK model.  The NONMEM
estimation model (`pk_busulfan_ucsf_2025.mod`, ADVAN3 TRANS4) incorporates:

* Allometric fat-free mass (FFM) scaling of clearance (estimated allometric
  exponent) and volume
* Sex effect on volume
* Maturation function for paediatric CL
* Time-varying clearance (decreasing exponentially with time post-dose)
* Between-subject variability (IIV): `BLOCK(2)` on CL and V1 (with
  covariance), separate diagonal term for V2
* Inter-occasion variability (IOV) on CL (CV 11.6\%) and V (CV 11.2\%)
  across 5 dosing occasions

**100 patients** (80 paediatric 1--18 yr, 20 adult 18--70 yr) each received
4 doses at 24-hour intervals.  Doses 2--4 were adapted via **MAP Bayesian
estimation** (`PKPDmap::get_map_estimates()`) to target a cumulative AUC of
90 mg\*h/L.  Sampling occurred at 1, 2, 3, and 6 hours after the end of
the infusion (4, 5, 6, and 9 hours post-dose) for the first three doses
(12 observations per patient, 1200 total).

Proportional (6.9\%) and additive (28.6 ng/mL) residual error was added to
all simulated concentrations.

**Intentional errors** were then introduced and tracked via `FLAG` (integer
bitmask) and `FLAG_LABEL` (character description) columns:

| FLAG bit | Error type | Count |
|---|---|---|
| 1 | Covariate weight: decimal-place shift (e.g. 27.0 kg recorded as 2.7 kg) | 1 patient |
| 2 | Concentration IQR outlier (\>3\*IQR above Q3 per TAD bin) | 2 observations |
| 4 | Timing error (+/- 5, 10, 30, or 60 minutes from true time) | 10 observations |
| 8 | Concentration magnitude error (multipliers 0.01--2.0x) | 10 observations |

The full simulation script is in `data-raw/simulate_busulfan.R`.

---

## How the dataset was generated

The dataset was simulated using `PKPDsim` (>= 1.5.0) and `PKPDmap` (>= 1.1.6)
with the two-compartment busulfan PK model described above.  One hundred
patients were generated: 80 paediatric (age 1--18 years, weight drawn from a
log-normal distribution centred on an age-based mean) and 20 adult (age
18--70 years, weight ~N(72, 16) kg).  Between-subject variability was sampled
from the 3 x 3 IIV omega matrix (CL, V1, V2), and inter-occasion variability
was applied independently for each of the four 24-hour dosing occasions.

Each patient received four once-daily 3-hour intravenous infusions.  After
each dose, four TDM samples were collected at 4, 5, 6, and 9 hours post-dose
(i.e. 1, 2, 3, and 6 hours after the end of the infusion).  Doses 2--4 were
adapted using MAP Bayesian estimation (via `PKPDmap`) targeting a cumulative
AUC of 90,000 ng/mL*h (22,500 ng/mL*h per dose).

The full simulation script, including covariate generation, IOV sampling, MAP
dose adaptation, and deliberate error injection, is available in
`data-raw/simulate_busulfan.R`.

---

## 1. Load the dataset

```{r load-data}
# Load the bundled simulated dataset
dat <- read.csv(system.file("extdata", "busulfan_sim.csv", package = "irxclean"))

cat(sprintf(
  "Patients: %d  |  Total rows: %d  |  Observations: %d\n",
  length(unique(dat$ID)), nrow(dat), sum(dat$EVID == 0)
))
head(dat[dat$ID == 1, ], 6)
```

```{r summarise-errors}
# Errors introduced in the simulation
err_obs <- dat[
  dat$FLAG > 0 & dat$EVID == 0,
  c("ID", "TIME", "DV", "FLAG", "FLAG_LABEL")
]
cat("Flagged observations:\n")
print(err_obs[order(err_obs$FLAG, err_obs$ID), ])
```

---

## 1b. Explore: concentration vs. time-after-dose

```{r plot-pk-setup, include=FALSE}
obs <- dat[dat$EVID == 0, ]
obs$TAD <- obs$TIME %% 24

obs$error_type <- "Clean"
obs$error_type[grepl("magnitude", obs$FLAG_LABEL)] <- "Magnitude error"
obs$error_type[grepl("timing",    obs$FLAG_LABEL)] <- "Timing error"
obs$error_type[grepl("conc_iqr",  obs$FLAG_LABEL)] <- "Conc. outlier"
obs$error_type[grepl("covariate", obs$FLAG_LABEL)] <- "Covariate error"
obs$error_type <- factor(obs$error_type,
  levels = c("Clean", "Covariate error", "Conc. outlier",
             "Timing error", "Magnitude error"))

pal <- c(
  "Clean"           = "#AAAAAA",
  "Covariate error" = "#00769E",
  "Conc. outlier"   = "#AD0000",
  "Timing error"    = "#657422",
  "Magnitude error" = "#E07800"
)
sz <- c(
  "Clean"           = 0.7,
  "Covariate error" = 2.0,
  "Conc. outlier"   = 2.5,
  "Timing error"    = 2.0,
  "Magnitude error" = 2.0
)

obs_clean <- obs[obs$error_type == "Clean", ]
obs_err   <- obs[obs$error_type != "Clean", ]

.pk_layers <- function(x_var, x_lab, x_breaks) {
  list(
    ggplot2::geom_point(
      data   = obs_clean,
      ggplot2::aes(x = .data[[x_var]], y = DV),
      colour = pal["Clean"], size = sz["Clean"], alpha = 0.25
    ),
    ggplot2::geom_point(
      data  = obs_err,
      ggplot2::aes(x = .data[[x_var]], y = DV,
                   colour = error_type, size = error_type),
      alpha = 0.9
    ),
    ggplot2::scale_colour_manual(values = pal, drop = FALSE,
                                 name = "Error type"),
    ggplot2::scale_size_manual(values = sz, drop = FALSE, guide = "none"),
    ggplot2::scale_y_log10(labels = scales::label_comma(),
                           breaks = c(1, 10, 100, 1000, 10000)),
    ggplot2::scale_x_continuous(breaks = x_breaks),
    ggplot2::xlab(x_lab),
    ggplot2::ylab("DV (ng/mL, log scale)"),
    ggplot2::theme_bw(),
    ggplot2::theme(legend.position  = "right",
                   panel.grid.minor = ggplot2::element_blank())
  )
}
```

```{r plot-pk-tad, fig.width=8, fig.height=5, fig.cap="DV vs. time after dose (TAD). Observations overlay across all four dosing occasions. Sampling at 4, 5, 6, and 9 h post-dose (1, 2, 3, 6 h after end of 3-h infusion)."}
ggplot2::ggplot() +
  .pk_layers("TAD", "Time after dose (h)", c(4, 5, 6, 9)) +
  ggplot2::labs(
    title    = "DV vs. time after dose",
    subtitle = sprintf("%d patients | %d obs | %d flagged",
                       dplyr::n_distinct(obs$ID), nrow(obs),
                       sum(obs$error_type != "Clean"))
  )
```

```{r plot-pk-time, fig.width=9, fig.height=5, fig.cap="DV vs. absolute time. Vertical dashed lines mark the four dose times (0, 24, 48, 72 h). The within-occasion PK shape is visible at each occasion; the covariate-error patient shows a shifted curve across all occasions."}
dose_times <- unique(dat$TIME[dat$EVID == 1])

ggplot2::ggplot() +
  .pk_layers("TIME", "Time (h)", c(0, 24, 48, 72, 81)) +
  ggplot2::geom_vline(xintercept = dose_times, linetype = "dashed",
                      colour = "#999999", linewidth = 0.4) +
  ggplot2::labs(
    title    = "DV vs. absolute time",
    subtitle = sprintf("%d patients | %d obs | %d flagged",
                       dplyr::n_distinct(obs$ID), nrow(obs),
                       sum(obs$error_type != "Clean"))
  )
```

---

## 2. Gross data quality assessment

Before running NONMEM, `assess_data_quality()` performs rapid model-independent
screening: covariate outliers, concentration elevations, TAD-bin outliers, and
missing data.

```{r gross-cleaning}
qc <- assess_data_quality(
  data             = dat,
  covariate_cols   = c("AGE", "WT", "HT"),
  categorical_cols = "SEX",
  iqr_multiplier   = 3,
  conc_increase_threshold = 1.5,
  conc_floor       = 1    # busulfan LLOQ ~1 ng/mL; skip ratio check below this
)
```

```{r gross-covariate-outliers}
cat("Covariate outliers (note weight decimal error):\n")
print(qc$covariate_outliers)
```

```{r gross-conc-outliers}
cat("Concentration TAD-bin outliers:\n")
print(qc$concentration_bin_outliers)
```

```{r gross-missing}
n_missing <- qc$missing_summary[qc$missing_summary$n_missing > 0, ]
if (nrow(n_missing) == 0) {
  cat("Missing data: none\n")
} else {
  print(n_missing)
}
```

The gross screen flags the two IQR-outlier concentrations and the weight
decimal error.

---

## 3. Iterative observation flagging (NONMEM required)

`remove_erroneous_obs()` runs NONMEM iteratively, removing the observation with
the largest |CWRES| at each step and refitting.  Results are cached to
`busulfan_results.RDS` in the vignette working directory so subsequent renders
skip the NONMEM run.  If neither a cache file nor a PsN/NONMEM installation is
available, the 5-iteration testdata bundled with the package is used as a
fallback.

```{r remove-obs}
cache_file <- "busulfan_results.RDS"

if (file.exists(cache_file)) {
  results <- readRDS(cache_file)
  message("Loaded cached results from ", cache_file)
} else if (nzchar(Sys.which("execute"))) {
  file.copy(
    system.file("extdata", "pk_busulfan_ucsf_2025.mod", package = "irxclean"),
    "pk_busulfan_ucsf_2025.mod", overwrite = TRUE
  )
  file.copy(
    system.file("extdata", "busulfan_sim.csv", package = "irxclean"),
    "busulfan_sim.csv", overwrite = TRUE
  )
  results <- remove_erroneous_obs(
    dat    = "busulfan_sim",
    mod    = "pk_busulfan_ucsf_2025",
    run_id = "busulfan_removal",
    n      = 20
  )
  saveRDS(results, cache_file)
  message("Results saved to ", cache_file)
} else {
  message("NONMEM/PsN not found and no cache -- loading package testdata.")
  results <- readRDS(system.file("testdata", "rem_test_results.RDS",
                                 package = "irxclean"))
}

cat("Iterations completed:", nrow(results$rmse), "\n")
cat("Flagged observations (ranked by |CWRES|):\n")
print(results$rem[, c("ID", "TIME", "DV", "CWRES")])
```

---

## 4. Removal metrics

`plot_removal_metrics()` visualises how model fit statistics evolve as
successive observations are removed.

```{r metrics-rmse, fig.cap="RMSE across iterations"}
plot_removal_metrics(results, metric = "rmse")
```

```{r metrics-pofv, fig.cap="pOFV across iterations"}
plot_removal_metrics(results, metric = "pOFV")
```

```{r metrics-thetas, fig.cap="Theta estimates across iterations"}
plot_removal_metrics(results, metric = "thetas")
```

---

## 5. Model stability check

`check_model_stability()` compares parameter estimates at each iteration
against the baseline and flags iterations where any theta drifts beyond a
threshold percentage.

```{r stability}
stability <- check_model_stability(results, threshold_pct = 20)
print(stability)
```

---

## 6. Demographic comparison

`plot_demographic_comparison()` tests whether subjects with flagged
observations differ demographically from the rest of the cohort.

```{r demographics, fig.cap="Demographic comparison: removed vs retained"}
plot_demographic_comparison(
  data             = dat,
  results          = results,
  continuous_cols  = c("AGE", "WT", "HT"),
  categorical_cols = "SEX"
)
```

---

## 7. Apply final exclusions

After reviewing the ranked flagged observations, `remove_observations_from_data()`
applies the user-approved exclusions to produce the analysis-ready dataset.

```{r remove-from-data, eval=FALSE}
cleaned <- remove_observations_from_data(
  data        = dat,
  results     = results,
  n_to_remove = 3
)

write.csv(cleaned, "busulfan_cleaned.csv", row.names = FALSE)
cat("Cleaned dataset:", nrow(cleaned), "rows\n")
```

---

## 8. Generate report

A single call to `generate_report()` compiles an HTML summary of all findings.

```{r report, eval=FALSE}
generate_report(
  results              = results,
  data                 = dat,
  quality_results      = qc,
  output_file          = "busulfan_irxclean_report.html",
  continuous_cov_cols  = c("AGE", "WT", "HT"),
  categorical_cov_cols = "SEX"
)
```

---

## Session information

```{r session-info}
sessionInfo()
```
