---
title: "Simulated Busulfan PK: Part 2 (ferx) -- Model-Informed Iterative Cleaning"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulated Busulfan PK: Part 2 (ferx) -- Model-Informed Iterative Cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5
)
library(irxclean)
```

## Overview

This vignette demonstrates the **ferx engine** for model-informed iterative
observation cleaning.  It mirrors the NONMEM-based workflow in Part 2
(`busulfan_pt2_modelbased_cleaning`) but uses `engine = "ferx"` instead of
NONMEM/PsN, requiring only the R package
[ferx](https://github.com/InsightRX/ferx) -- no external tools.

> **Prerequisites:** This vignette continues from **Part 1**
> (`busulfan_pt1_gc_exclusion`).  Run Part 1 first and save
> `busulfan_pt1_ready.csv` to your working directory before proceeding.

```{r check-ferx, include=FALSE}
has_ferx <- requireNamespace("ferx", quietly = TRUE)
```

---

## 1. Load the Phase 1 dataset

```{r load-data}
pt1_file <- "busulfan_pt1_ready.csv"

if (file.exists(pt1_file)) {
  dat <- read.csv(pt1_file)
  message("Loaded Phase 1 output: ", pt1_file)
} else {
  message(
    "busulfan_pt1_ready.csv not found -- falling back to raw dataset.\n",
    "Run busulfan_pt1_gc_exclusion first to generate the cleaned input."
  )
  dat <- read.csv(system.file("extdata", "busulfan_sim.csv",
                               package = "irxclean"))
}

cat(sprintf(
  "Patients: %d  |  Rows: %d  |  Observations: %d\n",
  length(unique(dat$ID)), nrow(dat), sum(dat$EVID == 0)
))
```

---

## 2. Define the ferx model

A `.ferx` model file describes the structural PK model, random effects, and
error model.  For this busulfan dataset we use a one-compartment IV infusion
model with IIV on CL and V and proportional residual error.

```{r write-model}
model_text <- '
[parameters]
  theta TVCL(4.0, 0.1, 100.0)  ; Clearance
  theta TVV(15.0, 1.0, 500.0)  ; Volume

  omega ETA_CL ~ 0.09  ; IIV CL
  omega ETA_V  ~ 0.04  ; IIV V

  sigma PROP_ERR ~ 0.1 (sd)  ; Proportional error

[individual_parameters]
  CL = TVCL * exp(ETA_CL)
  V  = TVV  * exp(ETA_V)

[structural_model]
  ode(states=[central])

[odes]
  d/dt(central) = -(CL/V) * central

[scaling]
  y = central / V

[error_model]
  DV ~ proportional(PROP_ERR)

[fit_options]
  method  = focei
  maxiter = 300
'

model_path <- file.path(tempdir(), "busulfan_1cpt.ferx")
writeLines(model_text, model_path)
cat("Model written to:", model_path, "\n")
```

---

## 3. Iterative observation flagging with ferx

`remove_erroneous_obs()` with `engine = "ferx"` calls `ferx::ferx_fit()`
at each iteration instead of NONMEM.  The same algorithm applies: identify
the observation with the largest |CWRES|, remove it, and re-fit.

Results are cached to an RDS file so subsequent renders skip the fitting.

```{r remove-obs, eval=FALSE}
cache_file <- "busulfan_ferx_results.RDS"

if (file.exists(cache_file)) {
  results <- readRDS(cache_file)
  message("Loaded cached results from ", cache_file)
} else {
  results <- remove_erroneous_obs(
    dat             = dat,
    mod             = model_path,
    run_id          = "bu_ferx",
    n               = 20,
    verbose         = TRUE,
    save_results    = FALSE,
    stability_check = TRUE,
    engine          = "ferx",
    method          = "focei"
  )
  saveRDS(results, cache_file)
  message("Results saved to ", cache_file)
}

cat("Iterations completed:", nrow(results$rmse), "\n")
cat("Flagged observations (ranked by |CWRES|):\n")
print(results$rem[, c("ID", "TIME", "DV", "CWRES")])
```

```{r fallback, echo=FALSE}
message("ferx package not installed -- loading pre-computed NONMEM results as fallback.")
results <- readRDS(system.file("extdata", "busulfan_results.RDS",
                                package = "irxclean"))
```

---

## 4. Removal metrics

The same plotting functions work identically regardless of engine.

```{r metrics-pofv, fig.cap="Change in pseudo-OFV per iteration."}
plot_removal_metrics(results, metric = "pOFV", verbose = FALSE)
```

```{r metrics-thetas, fig.cap="THETA estimates normalised to final value."}
plot_removal_metrics(results, metric = "thetas", verbose = FALSE)
```

```{r metrics-omegas, fig.cap="OMEGA estimates normalised to final value."}
plot_removal_metrics(results, metric = "omegas", verbose = FALSE)
```

```{r metrics-sigmas, fig.cap="SIGMA estimates normalised to final value."}
plot_removal_metrics(results, metric = "sigmas", verbose = FALSE)
```

```{r metrics-rmse, fig.cap="nRMSE across iterations."}
plot_removal_metrics(results, metric = "rmse", verbose = FALSE)
```

---

## 5. Model stability

```{r stability}
stability <- check_model_stability(results, threshold_pct = 20)
print(stability)
```

```{r stability-flags}
flag_tbl <- stability_flag_table(stability)
if (nrow(flag_tbl) > 0) knitr::kable(flag_tbl, digits = 2) else
  cat("All parameters stable.\n")
```

---

## 6. Demographic comparison

```{r demographics, fig.width=10, fig.height=7, fig.cap="Demographic comparison: subjects with removed observations vs. the rest of the cohort."}
plot_demographic_comparison(
  data             = dat,
  results          = results,
  continuous_cols  = c("AGE", "WT", "HT"),
  categorical_cols = "SEX"
)
```

---

## 7. Apply final exclusions

After reviewing the flagged observations, apply user-approved exclusions:

```{r remove-from-data, eval=FALSE}
cleaned <- remove_observations_from_data(
  data    = dat,
  results = results,
  n       = 10
)

write.csv(cleaned, "busulfan_ferx_cleaned.csv", row.names = FALSE)
cat("Cleaned dataset:", nrow(cleaned), "rows\n")
```

---

## 8. Generate HTML report

```{r report, eval=FALSE}
generate_report(
  results              = results,
  data                 = dat,
  continuous_cov_cols  = c("AGE", "WT", "HT"),
  categorical_cov_cols = "SEX",
  title                = "Busulfan PK -- ferx Model-Informed Cleaning Report",
  output_file          = "busulfan_ferx_report.html"
)
```

---

## Session information

```{r session-info}
sessionInfo()
```
