---
title: "Simulated Busulfan PK: Part 1 — Gross QC and Exclusion Criteria"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulated Busulfan PK: Part 1 — Gross QC and Exclusion Criteria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5
)
library(irxclean)
```

## Overview

This vignette covers **Phase 1** of an `irxclean` analysis: model-independent data
screening before a base NONMEM model is fitted.  It walks through:

- Loading and visualising the simulated busulfan dataset
- Gross data quality screening with `assess_data_quality()`
- Investigating covariate anomalies that IQR-based screening misses
- Applying structured exclusion criteria with `apply_exclusion_criteria()`
- Saving the cleaned, analysis-ready dataset for **Part 2**

> **Part 2** (`busulfan_pt2_modelbased_cleaning`) picks up from the output of this
> vignette and demonstrates NONMEM-based iterative CWRES outlier removal.  Run this
> vignette first and save `busulfan_pt1_ready.csv` before proceeding.

### Dataset and intentional errors

The dataset (`busulfan_sim.csv`) was generated with **PKPDsim** using a two-compartment
IV busulfan population PK model (ADVAN3 TRANS4, allometric FFM scaling, maturation,
time-varying CL, IIV + IOV).  100 patients (80 paediatric, 20 adult) each received 4
MAP-adapted doses, with 4 TDM samples per dose (1,200 observations total).

The following intentional errors were introduced:

| FLAG bit | Error type | Count |
|---|---|---|
| 1 | Covariate weight: decimal-place shift (e.g. 22.0 kg -> 2.2 kg) | 1 patient |
| 2 | Concentration IQR outlier (>3\*IQR above Q3 per TAD bin) | 2 observations |
| 4 | Timing error (+/- 5, 10, 30, or 60 minutes from true time) | 10 observations |
| 8 | Concentration magnitude error (multipliers 0.25--1.75x) | 10 observations |

---

## 1. Load the dataset

```{r load-data}
dat <- busulfan_sim

cat(sprintf(
  "Patients: %d  |  Rows: %d  |  Observations: %d\n",
  length(unique(dat$ID)), nrow(dat), sum(dat$EVID == 0)
))
```

```{r summarise-errors}
err_obs <- dat[dat$FLAG > 0 & dat$EVID == 0,
               c("ID", "TIME", "DV", "FLAG", "FLAG_LABEL")]
cat("Flagged observations (ground truth):\n")
print(err_obs[order(err_obs$FLAG, err_obs$ID), ])
```

The `FLAG` and `FLAG_LABEL` columns record the ground-truth errors introduced during
simulation.  In a real analysis these columns would not exist; they are shown here to
let you assess how well each screening method performs.

---

## 2. Concentration vs. time-after-dose

Visualising the raw concentration data is the first step in any PK cleaning workflow.
Plotting DV against both time-after-dose (TAD) and absolute time lets you spot
structurally implausible values before any model is fitted.

```{r plot-pk-setup, include=FALSE}
obs <- dat[dat$EVID == 0, ]
obs$TAD <- obs$TIME %% 24

obs$error_type <- "Clean"
obs$error_type[grepl("magnitude", obs$FLAG_LABEL)] <- "Magnitude error"
obs$error_type[grepl("timing",    obs$FLAG_LABEL)] <- "Timing error"
obs$error_type[grepl("conc_iqr",  obs$FLAG_LABEL)] <- "Conc. outlier"
obs$error_type[grepl("covariate", obs$FLAG_LABEL)] <- "Covariate error"
obs$error_type <- factor(obs$error_type,
  levels = c("Clean", "Covariate error", "Conc. outlier",
             "Timing error", "Magnitude error"))

pal <- c(
  "Clean"           = "#AAAAAA",
  "Covariate error" = "#00769E",
  "Conc. outlier"   = "#AD0000",
  "Timing error"    = "#657422",
  "Magnitude error" = "#E07800"
)
sz <- c(
  "Clean"           = 0.7,
  "Covariate error" = 2.0,
  "Conc. outlier"   = 2.5,
  "Timing error"    = 2.0,
  "Magnitude error" = 2.0
)

obs_clean <- obs[obs$error_type == "Clean", ]
obs_err   <- obs[obs$error_type != "Clean", ]

.pk_layers <- function(x_var, x_lab, x_breaks) {
  list(
    ggplot2::geom_point(
      data   = obs_clean,
      ggplot2::aes(x = .data[[x_var]], y = DV),
      colour = pal["Clean"], size = sz["Clean"], alpha = 0.25
    ),
    ggplot2::geom_point(
      data  = obs_err,
      ggplot2::aes(x = .data[[x_var]], y = DV,
                   colour = error_type, size = error_type),
      alpha = 0.9
    ),
    ggplot2::scale_colour_manual(values = pal, drop = FALSE,
                                 name = "Error type"),
    ggplot2::scale_size_manual(values = sz, drop = FALSE, guide = "none"),
    #ggplot2::scale_y_log10(labels = scales::label_comma(),
                          # breaks = c(1, 10, 100, 1000, 10000)),
    ggplot2::scale_x_continuous(breaks = x_breaks),
    ggplot2::xlab(x_lab),
    ggplot2::ylab("DV (ng/mL, log scale)"),
    ggplot2::theme_bw(),
    ggplot2::theme(legend.position  = "right",
                   panel.grid.minor = ggplot2::element_blank())
  )
}
```

```{r plot-pk-tad, fig.width=8, fig.height=5, fig.cap="DV vs. time after dose (TAD). TDM samples drawn randomly from four post-infusion windows (10-45 min, 1-3 h, 3-6 h, 6-12 h after end of the 3-h infusion)."}
ggplot2::ggplot() +
  .pk_layers("TAD", "Time after dose (h)", c(3, 6, 9, 12, 15)) +
  ggplot2::labs(
    title    = "DV vs. time after dose",
    subtitle = sprintf("%d patients | %d obs | %d flagged",
                       dplyr::n_distinct(obs$ID), nrow(obs),
                       sum(obs$error_type != "Clean"))
  )
```

```{r plot-pk-time, fig.width=9, fig.height=5, fig.cap="DV vs. absolute time. Vertical dashed lines mark the four dose times (0, 24, 48, 72 h)."}
dose_times <- unique(dat$TIME[dat$EVID == 1])

ggplot2::ggplot() +
  .pk_layers("TIME", "Time (h)", c(0, 24, 48, 63, 72)) +
  ggplot2::geom_vline(xintercept = dose_times, linetype = "dashed",
                      colour = "#999999", linewidth = 0.4) +
  ggplot2::labs(
    title    = "DV vs. absolute time",
    subtitle = sprintf("%d patients | %d obs | %d flagged",
                       dplyr::n_distinct(obs$ID), nrow(obs),
                       sum(obs$error_type != "Clean"))
  )
```

The plots use ground-truth labels for illustration.  In practice you would look for
points that stand out visually — concentrations far above or below the main cloud,
unexpected rises without preceding doses, or clusters of observations whose shape
differs markedly from the rest of the cohort.

---

## 3. Gross data quality assessment

`assess_data_quality()` performs rapid, model-independent screening across four
dimensions: covariate outliers (IQR-based), concentration TAD-bin outliers, unexplained
concentration elevations, and missing data.

```{r gross-qc}
qc <- assess_data_quality(
  data             = dat,
  covariate_cols   = c("AGE", "WT", "HT"),
  categorical_cols = "SEX",
  iqr_multiplier   = 3,
  conc_increase_threshold = 1.5,
  conc_floor       = 1
)
```

### 3.1 Covariate outliers

```{r qc-covariate-outliers}
cat("Covariate outliers detected by IQR screen:\n")
print(qc$covariate_outliers)
```

The IQR screen identifies several adult subjects (high AGE) who stand out in a
predominantly paediatric cohort — a real finding that warrants a protocol check.

Notice that **the weight decimal error (ID 18, WT = 2.2 kg) is not flagged here.**
With 80 paediatric patients in the cohort, the lower IQR bound for WT is clamped to
zero (Q1 - 3 x IQR < 0), so a value of 2.2 kg passes the test.  Section 4 shows how
to detect this type of error through direct inspection and physiology-guided thresholds.

### 3.2 Concentration TAD-bin outliers

```{r qc-conc-bin-outliers}
cat("Concentration TAD-bin outliers:\n")
print(qc$concentration_bin_outliers)
```

The IQR screen catches the two grossly elevated concentrations introduced as FLAG=2
errors, along with any other observations that fall outside the distribution for their
TAD window.  This provides a first-pass list for review; the model-based approach in
Part 2 will refine this further using CWRES.

### 3.3 Concentration elevation flags

```{r qc-conc-flags}
cat("Concentration elevations without intervening dose:\n")
print(qc$concentration_flags)
```

### 3.4 Missing data summary

```{r qc-missing}
cat("Columns with missing values:\n")
miss <- qc$missing_summary[qc$missing_summary$n_missing > 0, ]
if (nrow(miss) == 0) cat("None.\n") else print(miss)
```

### 3.5 QC plot

```{r qc-plot, fig.width=9, fig.height=6}
plot(qc)
```

---

## 4. Investigating the weight decimal error

As noted above, a single patient has an implausibly low body weight due to a
decimal-place transcription error (22.0 kg entered as 2.2 kg).  The IQR rule does not
catch this because the lower bound for the paediatric-skewed WT distribution is clamped
to zero.

Direct range inspection reveals the anomaly immediately:

```{r wt-range}
# Per-subject weight (one row per patient)
subj <- dat[!duplicated(dat$ID), c("ID", "AGE", "WT", "HT")]
cat("WT range across subjects:\n")
print(range(subj$WT))

cat("\nSubjects with WT < 5 kg (physiologically implausible for this study):\n")
print(subj[subj$WT < 5, ])
```

A busulfan protocol treating patients aged 0--70 years would never have a patient
weighing 4 kg; the minimum physiologically plausible weight in this context is
approximately 5 kg.  The simple threshold `WT < 5` cleanly identifies ID 76.

Before applying the exclusion, confirm the error by reviewing raw dose records for
the subject:

```{r wt-error-subject}
cat("All rows for ID 18:\n")
print(dat[dat$ID == 76, c("ID", "TIME", "AMT", "EVID", "AGE", "WT", "HT", "FLAG_LABEL")])
```

The dosing was calculated using the erroneous weight, so both the covariate and the
dose history are unreliable.  Excluding the subject entirely (pending data source
verification) is the safest course of action.

---

## 5. Apply exclusion criteria

`apply_exclusion_criteria()` applies a standardised battery of model-independent
checks.  The `custom_criteria` argument allows user-defined logical vectors to be
included alongside the built-in checks.

Here we add one custom criterion — `wt_implausible` — that flags all rows for ID 18:

```{r exclusion-criteria}
excl <- apply_exclusion_criteria(
  data             = dat,
  check_dose_outlier       = TRUE,
  dose_iqr_multiplier      = 3,
  conc_increase_threshold  = 1.5,
  conc_floor               = 1,
  custom_criteria  = list(
    wt_implausible = dat$ID == 76   # decimal-place weight error
  )
)
```

```{r exclusion-summary}
print(excl)
```

```{r exclusion-table}
if (nrow(excl$exclusion_summary) > 0) {
  knitr::kable(excl$exclusion_summary, digits = 2,
               caption = "Exclusion criteria summary")
}
```

The `excl$data_clean` object contains the dataset after all flagged rows have been
removed.

---

## 6. Summary

```{r summary}
n_raw     <- nrow(dat)
n_clean   <- nrow(excl$data_clean)
n_removed <- n_raw - n_clean

subj_raw   <- length(unique(dat$ID))
subj_clean <- length(unique(excl$data_clean$ID))

cat(sprintf(
  "Raw dataset  : %d rows, %d subjects\n",
  n_raw, subj_raw
))
cat(sprintf(
  "After Phase 1: %d rows, %d subjects (%d rows removed)\n",
  n_clean, subj_clean, n_removed
))
```

**Phase 1 findings:**

- **Covariate anomaly (ID 18):** Weight decimal-place error (4.0 kg; true ~40 kg).
  Subject excluded entirely pending source data verification.
- **Concentration IQR outliers:** Flagged by `assess_data_quality()`; passed forward
  for model-based review in Part 2.
- **Exclusion criteria:** No subjects lacked dose records or TDM observations.
  No samples were collected during an ongoing infusion.  One subject excluded via
  custom covariate criterion.

The `excl$data_flagged` data frame retains all rows with `excl_*` flag columns
appended — useful for audit trails and downstream documentation.

---

## 7. Save the analysis-ready dataset

Write the cleaned dataset to disk for use in Part 2:

```{r save-output, eval=FALSE}
write.csv(excl$data_clean, "busulfan_pt1_ready.csv", row.names = FALSE)
cat("Saved:", nrow(excl$data_clean), "rows to busulfan_pt1_ready.csv\n")
```

> **Next step:** Open `busulfan_pt2_modelbased_cleaning` to continue with
> NONMEM-based iterative CWRES outlier removal using the dataset saved above.

---

## Session information

```{r session-info}
sessionInfo()
```
