---
title: "Simulated Busulfan PK: Part 2 — Model-Informed Iterative Cleaning"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulated Busulfan PK: Part 2 — Model-Informed Iterative Cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5
)
library(irxclean)
```

## Overview

> **Prerequisites:** This vignette continues from **Part 1**
> (`busulfan_pt1_gc_exclusion`).  Run Part 1 first and save
> `busulfan_pt1_ready.csv` to your working directory before proceeding here.
> Part 1 covers dataset loading, concentration visualisation, gross QC with
> `assess_data_quality()`, covariate anomaly investigation, and structured
> exclusion criteria.

This vignette walks through the **model-informed** cleaning stages of an
`irxclean` analysis using the simulated busulfan pharmacokinetic dataset.  It
covers Phase 2 of the full workflow: NONMEM-based iterative observation
flagging, parameter stability assessment, demographic comparison, and HTML
report generation.

### Intentional errors remaining after Phase 1

After Part 1 removes the subject with a weight decimal error (ID 18), the
following errors remain in the dataset for the model to detect:

| FLAG bit | Error type | Count |
|---|---|---|
| 2 | Concentration IQR outlier (>3\*IQR above Q3 per TAD bin) | 2 observations |
| 4 | Timing error (+/- 5, 10, 30, or 60 minutes from true time) | 10 observations |
| 8 | Concentration magnitude error (multipliers 0.25--1.75x) | 10 observations |

---

## 1. Load the Phase 1 dataset

Load `busulfan_pt1_ready.csv` saved at the end of Part 1.  If that file is
not present in the working directory (e.g. when rendering from the package
without running Part 1 first), the raw simulated dataset is used as a
fallback — note that in this case one extra subject (ID 18) with a covariate
error remains in the data.

```{r load-data}
pt1_file <- "busulfan_pt1_ready.csv"

if (file.exists(pt1_file)) {
  dat <- read.csv(pt1_file)
  message("Loaded Phase 1 output: ", pt1_file)
} else {
  message(
    "busulfan_pt1_ready.csv not found -- falling back to raw dataset.\n",
    "Run busulfan_pt1_gc_exclusion first to generate the cleaned input."
  )
  dat <- busulfan_sim
}

cat(sprintf(
  "Patients: %d  |  Rows: %d  |  Observations: %d\n",
  length(unique(dat$ID)), nrow(dat), sum(dat$EVID == 0)
))
```

---

## 2. Iterative observation flagging (NONMEM required)

`remove_erroneous_obs()` fits NONMEM iteratively, removing the observation
with the largest |CWRES| at each step.  Three input modes are supported — see
`?remove_erroneous_obs` for details.  Here we demonstrate **Mode 3**: pass
the dataset as an in-memory R data frame (no separate file copy needed) and
point `mod` at the bundled model file path.

Results are cached to `busulfan_results.RDS` in the working directory so
subsequent renders skip the NONMEM run.  If neither a cache file nor a
PsN/NONMEM installation is present, the 5-iteration pre-computed result
bundled with the package is used as a fallback.

```{r remove-obs}
cache_file <- "busulfan_results.RDS"

if (file.exists(cache_file)) {
  results <- readRDS(cache_file)
  message("Loaded cached results from ", cache_file)
} else if (nzchar(Sys.which("execute"))) {
  # Mode 3: pass data frame directly; model given as a file path
  results <- remove_erroneous_obs(
    dat    = dat,   # Phase 1 cleaned data frame
    mod    = system.file("extdata", "pk_busulfan_ucsf_2025.mod",
                         package = "irxclean"),
    run_id = "busulfan_removal",
    n      = 20
  )
  saveRDS(results, cache_file)
  message("Results saved to ", cache_file)
} else {
  message("NONMEM/PsN not found and no cache -- loading package testdata.")
  results <- readRDS(system.file("extdata", "busulfan_results.RDS",
                                 package = "irxclean"))
}

cat("Iterations completed:", nrow(results$rmse), "\n")
cat("Flagged observations (ranked by |CWRES|):\n")
print(results$rem[, c("ID", "TIME", "DV", "CWRES")])
```

---

## 3. Removal metrics

```{r metrics-pofv, fig.cap="Change in pseudo-OFV per iteration. A sharp early drop followed by a plateau suggests meaningful early removals with diminishing returns."}
plot_removal_metrics(results, metric = "pOFV", verbose = FALSE)
```

```{r metrics-thetas, fig.cap="THETA estimates normalised to final value. Free axes per panel; phantom reference lines at 0.75 and 1.5 ensure consistent scale even when individual parameters are stable."}
plot_removal_metrics(results, metric = "thetas", verbose = FALSE)
```

```{r metrics-rmse, fig.cap="nRMSE across iterations."}
plot_removal_metrics(results, metric = "rmse", verbose = FALSE)
```

---

## 4. Error parameter trajectories

Proportional and additive residual error parameters are auto-detected from
the `param_labels` element of the results (parsed from `$THETA` comments in
the model file).  For this model, `THETA7` is proportional error and `THETA8`
is additive.  The detection source is printed when `quiet = FALSE` in
`generate_report()`.

```{r error-params}
detected <- irxclean:::.find_error_params(
  results$param_labels, results$par, verbose = TRUE
)
```

```{r traj-prop, fig.cap="Proportional residual error (THETA7) stability across iterations."}
irxclean:::.plot_param_trajectory(
  results,
  detected$prop_col,
  title = "Proportional residual error (THETA7) across iterations"
)
```

```{r traj-add, fig.cap="Additive residual error (THETA8) stability across iterations."}
irxclean:::.plot_param_trajectory(
  results,
  detected$add_col,
  title = "Additive residual error (THETA8) across iterations"
)
```

---

## 5. Model stability

```{r stability}
stability <- check_model_stability(results, threshold_pct = 20)
print(stability)
```

```{r stability-flags}
flag_tbl <- stability_flag_table(stability)
if (nrow(flag_tbl) > 0) knitr::kable(flag_tbl, digits = 2) else
  cat("All parameters stable.\n")
```

---

## 6. Demographic comparison

```{r demographics, fig.width=10, fig.height=7, fig.cap="Demographic comparison: subjects with removed observations vs. the rest of the cohort."}
plot_demographic_comparison(
  data             = dat,
  results          = results,
  continuous_cols  = c("AGE", "WT", "HT"),
  categorical_cols = "SEX"
)
```

---

## 7. Apply final exclusions

After reviewing the ranked flagged observations, apply user-approved exclusions:

```{r remove-from-data, eval=FALSE}
cleaned <- remove_observations_from_data(
  data    = dat,
  results = results,
  n       = 10  # remove top 10ranked observations
)

write.csv(cleaned, "busulfan_cleaned.csv", row.names = FALSE)
cat("Cleaned dataset:", nrow(cleaned), "rows\n")
```

---

## 8. Generate HTML report

`generate_report()` compiles all findings into a structured, self-contained
HTML report.  Proportional and additive error parameters are auto-detected
from `results$param_labels`; override explicitly if needed using either the
column name (`"THETA7"`) or NONMEM index notation (`"THETA(7)"`).

```{r report, eval=FALSE}
generate_report(
  results              = results,
  data                 = dat,
  # Auto-detected; shown here for illustration -- override if needed:
  prop_error_param     = "THETA(7)",
  add_error_param      = "THETA(8)",
  continuous_cov_cols  = c("AGE", "WT", "HT"),
  categorical_cov_cols = "SEX",
  title                = "Busulfan PK -- Model-Informed Cleaning Report",
  output_file          = "busulfan_irxclean_report.html",
  quiet                = FALSE   # prints auto-detection summary
)
```

---

## Session information

```{r session-info}
sessionInfo()
```
